Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

\abstract{%
Research on climate change impacts can require extensive processing, to
incorporate output from ensembles of different climate models and
different simulations of each model. This processing can be particularly
extensive when identifying and characterizing extreme events like heat
waves and frost day spells, as these must be processed from model output
with daily time steps and may use identification threholds based on
local distributions (e.g., 98th percentile temperature for that
location). Here, we provide an overview of working with daily climate
model output data in R. Further, we present the \pkg{futureheatwaves}
package, which we developed to ease the process of identifying,
characterizing, and exploring multi-day extreme events in climate model
output. This package takes as input a directory of projection files from
many climate models, identify all heat waves for multiple cities using a
user-specified heat wave definition, and write an output directory of
files with these heat waves and their characteristics (e.g., length,
absolute and relative temperatures). The \pkg{futureheatwaves} package
then allows the user to create custom functions and apply them across
all heat wave files; this functionality can be used either to explore
heat wave patterns (e.g., average length and intensity across different
climate models) or to apply epidemiological estimates to project health
impacts under different scenarios.
}

@@

The \pkg{futureheatwaves} package automates the process of identifying
and characterizing multi-day extreme events across different ensemble
members of one or more climate models. A variety of different heat wave
definitions have been used to identify heat waves in a time series of
temperature data \citep{smith2013heat}, and the choice of heat wave
definitions can influence both health effect estimates
\citep{chen2015influence, kent2014heat} and projected heat wave trends
\citep{smith2013heat}. Further, other types of extreme events will be
defined differently than heat waves (for example, frost day spells may
be defined as one or more days with temperature at or below \(0^oC\)).
This package therefore allows the user to create and use a custom the
extreme event definition used to identify events in the climate model
output. Finally, this package allows the user to explore the extreme
events identified in the climate model output with a function that can
take a user-defined R function and apply it across all extreme event
files generated for the separate ensemble members.

@@

The \pkg{futureheatwaves} package was facilitated to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). The extreme events are identified and
characterized at one or more study locations (e.g., cities), with study
locations specified by the user in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``). Second, the \code{apply\_all\_models} function
allows the user to apply custom functions across all the extreme event
data frames generated by \code{gen\_hw\_set} to summaries of extreme
events across all climate models and ensemble members (Figure
\ref{fig:overview}, right).

@@

The user can extensively customize the definition used to identify
extreme events. The default is a definition for heat waves (two or more
days with temperatures at or above the city's \(98^{th}\) percentile of
year-round temperature), but users can write a custom R function with
either a different heat wave definition or with a definition appropriate
for a different type of extreme event (e.g., one or more days at or
below \(0^oC\) for frost day spells). Some of these heat wave
characteristics are based on relative temperature, which is a measure of
how the temperature during a heat wave compares to the typical
distribution of temperature at that location \citep{IPCCch1}. These
relative temperature metrics will vary depending on whether you
calculate them based on a location's present-day temperature
distribution or on the location's temperature distribution in the
future, since typical temperature distributions in many locations are
expected to change with climate change. The package therefore allows the
user to specify date ranges of the temperature distributions to be used
in calculating these relative temperature metrics in each location.

Second, once the user creates these data frames of location-specific
heat waves, the \texttt{apply\_all\_models} function allows the user to
apply a custom function across all the heat wave data frames that were
generated to create a summary data frame (Figure \ref{fig:overview},
right). This functionality can be used to generate summary statistics
(e.g., determine average heat wave length or total heat wave days), or
can be used to apply more complex functions (e.g., apply epidemiologic
effect estimates across the heat waves to generate health impact
estimates).

@@

We have included data files in the package to serve as example files, so
users can try this package before applying it to their own directory of
climate projection files. The example files are all included as
comma-separated (.csv) files, rather than as saved R objects, because
the \texttt{gen\_hw\_set} function requires a directory of
comma-separated files as input and csv is a general file format familiar
to most users. This example data therefore also allow users to see how
to create the appropriate directory set-up for this package.

This example data comes from two climate models that are a part of
CMIP5: (1) the model of the Beijing Climate Center, China Meteorological
Administration (BCC) \citep{xin2013introduction} and (2) the National
Center for Atmospheric Research's (NCAR's) Community Climate System
Model, version 4 (CCSM4) \citep{gent2011community}. We include one
ensemble member from BCC (r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).
To ensure that the size of this example data is reasonably small, we
have only included projection data for grid points from these climate
models that are near five U.S. east coast cities: New York, NY;
Philadelphia, PA; Newark, NJ; Baltimore, MD, and Providence, RI.
Further, to keep the file sizes reasonably small, the historical
projections range over the years 1990 to 1999, while the future
projections are limited to 2060 to 2079. Users' applications of this
package will likely use directories with many more climate model
ensemble members and more locations; however, the operation of the
package is the same for this smaller example application as it would be
for a much larger application.

Once the \pkg{futureheatwaves} package is installed and loaded, the user
can find the local location of these files using R's
\texttt{system.file} function. In the later sections of this article, we
show how to use the package functions with these example files as
inputs.

When setting up to run climate model output data beyond this example
data, any downloaded CMIP5 files will require some pre-processing, in
terms of getting the getting the data into a specific format and storing
these files in a specific directory structure. A file of study locations
must also be saved locally in a specific format. Extensive details on
the required format of files and required input directory structure is
available in the \pkg{futureheatwaves} vignette ``futureheatwaves'',
while tips on processing CMIP5 files into the required structure are
given in the package's ``starting\_from\_netcdf'' vignette.

\subsection{Basic example of processing projection
files}\label{basic-example-of-processing-projection-files}

Once these files and directories are set up, \pkg{futureheatwaves} can
process them to identify and characterize heat waves in each ensemble
member's projection for each location using the \texttt{gen\_hw\_set}
function. For example, to process the example climate projections
directory included with the package, the user can run:

@@

This code first identifies and saves as objects the path names on the
user's computer of the example climate projections directory
(\texttt{projection\_dir\_location}) and the file of study locations
(\texttt{city\_file\_location}). The \texttt{gen\_hw\_set} function
processes this example input and creates a new directory,
\texttt{example\_results}, with files of identified and characterized
heat waves, in the user's current working directory. In this example
code, this processing is done using default values for the heat wave
definition, years for which to generate the heat wave data sets, etc.
Ways to customize these choices are fully explained later in the text.

In this call, the user must specify the directory where the results
should be written (\texttt{out}), the location of the directory of
climate projections (\texttt{dataFolder}), the names of the two main
subdirectories of that climate projection directory, as well as their
year boundaries (\texttt{dataDirectories}; see, for example, Figure
\ref{fig:directories}), the location of the city location file
(\texttt{citycsv}), and the names used for the grid coordinate, climate
projection, and projection date files (\texttt{coordinateFilenames},
\texttt{tasFilenames}, and \texttt{timeFilenames}). When
\texttt{gen\_hw\_set} is run, the user is advised that the function will
write files to his computer and must agree to proceed:

\begin{verbatim}
 Warning: This function will write new files to your computer in the 
 ~/tmp/example_results/ directory of your computer. If that directory already exists,
running this function will write over it. 
 Do you want to continue? (y / n): 
\end{verbatim}

\noindent The function provides status reports while processing. This
helps the user see that the function call is progressing, as
\texttt{gen\_hw\_set} can take a while to run if it is processing many
locations and / or many climate ensemble members. Once the function has
completed running, results will be written locally to the directory
specified by the \texttt{out} argument of \texttt{gen\_hw\_set}. The
structure of this output directory is shown in Figure
\ref{fig:directories} (right). This directory will include files with
some basic information about the climate models and the closest grid
points of each climate model to each location. The output directory will
also include a directory with files of identified and classified heat
waves for each ensemble member, including all characteristics in Table
\ref{tab:hwcharacteristics}.

@@

The default heat wave definition for this package is:

\begin{quote}
A \dfn{heat wave} is two or more days at or above a city-specific
threshold temperature, with the threshold determined as the \(98^{th}\)
percentile of year-round temperature in the city during some reference
period (by default, 1990--1999).
\end{quote}

Many different definitions of heat waves exist (e.g.,
\citet{smith2013heat, kent2014heat}), so researchers will often want to
use alternative ways to define heat waves. Researchers might want to use
a specific definition, for example, because it matches the definition
used by local health officials to declare heat wave warnings or, in the
case of health impact assessments, to match with a definition used in an
epidemiological study. Three components of the heat wave definition can
be easily customized in the \texttt{gen\_hw\_set} function call, without
creating a new R function to use to identify heat waves. The
customization of the heat wave definition is even more extensive as one
has the option of writing a custom R function.

First, the percentile used to identify a heat wave can be changed using
the \texttt{probThreshold} option in \texttt{gen\_hw\_set}. This option
can take values between 0 and 1. The default value is 0.98, or a
definition with a threshold of the \(98^{th}\) percentile of the
location's temperature. For example, to identify heat waves as two or
more days at or above the city's \(99^{th}\) percentile of year-round
temperature, the user could run:

\begin{Schunk}
\begin{Sinput}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           probThreshold = 0.99)
\end{Sinput}
\end{Schunk}

\noindent Second, the user can change the number of days used in the
heat wave definition using the \texttt{numDays} argument in the
\texttt{gen\_hw\_set} function. Combined, these two customization
choices can allow the user to identify heat waves using many of the heat
wave definitions used in previous climate and health research-- for
example, 9 of the 16 heat wave definitions used in \citet{kent2014heat}
could be fit using different combinations of these two options for
specifying threshold percentile and number of days.

Third, it is possible to specify the range of years that should be used
when determining this threshold. For example, if you wanted to base the
threshold for each location on its current climate, you could leave this
option as its default, while if you wanted to use a different set of
years (i.e.~a threshold relative to future projected temperatures), you
could set the start and end year bounds for this reference period using
the \texttt{thresholdBoundaries} argument in the function
\texttt{gen\_hw\_set}. For example, to use temperatures in each city
from 2070 to 2079 to determine threshold temperatures, you would run:

\begin{Schunk}
\begin{Sinput}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           thresholdBoundaries = c(2070, 2079))
\end{Sinput}
\end{Schunk}

It is also possible to use a completely customized heat wave definition.
This functionality allows the user to use heat wave definitions that
either require a number of days above an absolute threshold (e.g.,
maximum temperature of \(\ge95^oF\) for \(\ge1\) day;
\citet{kent2014heat, tan2007heat}) or that require a combination of
thresholds to be met (e.g., maximum daily temperature above a lower
threshold every day of the heat wave and above a higher threshold for a
certain number of days; \citet{kent2014heat, peng2011toward}).

To use a customized heat wave definition, the user can write and load an
R function that implements the definition. To work correctly, this
custom function must allow only specific inputs and generate only
specific outputs. Details about this required structure are provided in
the \pkg{futureheatwaves} package vignette.

To help in developing custom heat wave identification functions, the
package includes an example of the required input data frame,
\texttt{datafr}, which can be loaded with \texttt{data(datafr)}. If a
custom heat wave function is properly set up to use in this package, it
can take as input this example data frame and a threshold value and will
return the original data frame with the columns for whether the day was
part of a heat wave (\texttt{hw}) and, if so, the sequential number of
the heat wave out of all heat waves in that location
(\texttt{hw.number}).

The custom function can then be referenced using the
\texttt{IDheatwavesFunction} option in \texttt{gen\_hw\_set}, and it
will be used to identify heat waves in the data set. As an example, we
have included an alternative ID definition function in this package,
called \texttt{IDHeatwavesAlternative}, which identifies heat waves as
two or more days above the higher of either the location's \(98^{th}\)
percentile temperature or \(80^{o}F\), with the idea that this
definition prevents the identification of heat waves at mild
temperatures in locations with mild or cool climates. The following code
processes the example climate projecting files using this function as
the heat wave definition:

\begin{Schunk}
\begin{Sinput}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           IDheatwavesFunction = "IDHeatwavesAlternative")
\end{Sinput}
\end{Schunk}

With this ability to write and use a custom heat wave identification
function, most heat wave definitions can be applied with this package.
The exception is definitions that require two separate temperature
metrics (e.g., maximum temperature above a certain threshold or minimum
temperature above a different threshold;
\citet{kent2014heat, robinson2001definition}), because the
\texttt{gen\_hw\_set} function cannot take as input projections of more
than one temperature metric. Definitions that use a heat metric that
combines multiple inputs (e.g., heat index, which is a combined measure
of temperature and air moisture) can be used, but require that the
calculation of the metric is done when creating the input files to
include in the ``Climate projections" directory (Figure
\ref{fig:overview}, left), rather than as part of the heat wave
definition function.

To increase processing speed when identifying heat waves, we coded parts
of the default heat wave identification function in C++ and synced it
with R using the \CRANpkg{Rcpp} package \citep{Rcpp}. Users should
consider a similar strategy for custom heat wave definitions, especially
if processing a large number of climate projection files.

@@

Once you have created a directory of files with characterized heat waves
for each ensemble member, the results can be explored using the
\texttt{apply\_all\_models} function. This function allows the user to
apply custom R functions across all heat wave data frames created by the
\texttt{gen\_hw\_sets} call. The user can apply any R function that
follows certain standards in accepting input and returning output. Full
details on these standards are given in the \pkg{futureheatwaves}
package vignette.

As an example, if the user wanted to to get the average temperature of
the heat waves identified within each ensemble member, he or she could
write a simple function:

@@

\noindent The \texttt{apply\_all\_models} function can then apply this
\texttt{average\_mean\_temp} function across the heat wave data frames
for all ensemble members in all climate models:

@@

Note that the location of the directory with the heat wave data frames
must be specified using the \texttt{out} argument when calling
\code{apply\_all\_models}. Typically, this will be the directory path
for the directory specified with the \texttt{out} argument in
\code{gen\_hw\_set}. Location-specific results can be generated using
the \code{city\_specific} argument in \code{apply\_all\_models}:

@@

The \texttt{apply\_all\_models} function can also be used to project the
health impacts of heat waves. As a very simplistic example,
\citet{anderson2009weather} estimated that heat waves, defined as two or
more days at or above a community's \(98^{th}\) percentile temperature,
were associated with an added relative risk of 1.032 for
cardiorespiratory mortality risk in 107 U.S. communities. A simple
estimate of excess deaths associated with this added heat wave risk in a
community can be calculated as \citep{peng2011toward}:

\begin{equation}
E_c = N_c * (RR - 1) * L_c 
\label{eq:healthimpacts}
\end{equation}

@@

\begin{itemize}
\tightlist
\item
  \(E_c\) is the total number of excess deaths in community \(c\);
\item
  \(N_c\) is the baseline average daily mortality in community \(c\);
\item
  \(RR\) is the relative risk of cardiorespiratory mortality per day
  associated with a heat wave; and
\item
  \(L_c\) is the total number of heat wave days in community \(c\) over
  the study period.
\end{itemize}

This impact assessment equation can be translated into a function that
merges each projection's heat wave data frame with a data frame of
community-specific baseline mortality rates (\(B_c\)), calculates
equation \ref{eq:healthimpacts} for each heat wave, and then sums up the
total excess deaths across all heat waves:

\begin{verbatim}
excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
                                      by = "city") %>%
                dplyr::mutate(excess_deaths = base_mort * length * RR)
        out <- sum(hw_datafr$excess_deaths)
        return(out)
}
\end{verbatim}

\noindent Once defined in R, this function can be applied across all
heat waves from all climate models' ensemble members, provided that you
have a data frame called \texttt{city\_mortality} with columns with each
community's identifier (\texttt{city}) and baseline mortality rate
(\texttt{base\_mort}), using the call:

\begin{Schunk}
\begin{Sinput}
apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
\end{Sinput}
\end{Schunk}

As examples of the type of functions that can be used with
\texttt{apply\_all\_models}, we have included several simple functions
for exploring heat waves (Table \ref{tab:applyfunctions}). We included
an example data frame representative of the heat wave data frames that
\texttt{gen\_hw\_set} outputs, to use to test new functions. This data
can be loaded with \texttt{data(hw\_datafr)}. The process for creating a
new function to use to explore heatwaves should be to:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Load the example data frame with \texttt{load(hw\_datafr)};
\item
  Build a function that works with \texttt{hw\_datafr} as an input; and
\item
  Apply the function with \texttt{apply\_all\_models} to process all of
  the heat wave files.
\end{enumerate}

\begin{table}
\centering
\begin{tabular}{ll}
\toprule
Function & Description \\
\midrule
\code{number\_of\_heatwaves} & Determines the number of extreme events\\
\code{heatwave\_days} & Sums up the total number of extreme event days across all heat waves\\
\code{average\_length} & Calculates the average length of extreme events\\
\code{average\_mean\_temp} & Calculates the average mean temperature across all extreme events\\
\bottomrule
\end{tabular}
\caption{Example functions included in the \pkg{futureheatwaves} package that can be used to explore extreme events using the \code{apply\_all\_models} function.}
\label{tab:applyfunctions}
\end{table}

@@

The heat wave data sets characterize heat wave in several ways that are
based on relative temperature (for example, the
\texttt{mean.temp.quantile}, \texttt{max.temp.quantile}, and
\texttt{min.temp.quantile} columns described in Table
\ref{tab:hwcharacteristics}). These characteristics are measured for
each of the heat waves identified by the \texttt{gen\_hw\_set} function
by taking the absolute temperature of the heat wave (e.g., average
temperature during the heat wave is \(90^oF\)) and comparing it to the
heat wave location's typical temperature distribution. This process
generates relative measures of how intense the heat wave is compared to
what is normal in that location (e.g., \(90^oF\) is in the \(99^{th}\)
percentile of year-round temperatures in the location).

The values of these relative heat wave characteristics will depend on
what time period is used to determine the location's typical temperature
distribution. For example, the same heat wave would likely have much
higher values for these relative characteristics if the temperature
during the heat wave is compared to typical temperatures in the present
day for that location versus if the temperature is compared to typical
temperatures much later in the \(21^{st}\) century for the same
location.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
\end{center}
\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}. The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used. The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example.}
\label{fig:gridmap}
\end{figure}

>>>>

\abstract{%
Research on climate change impacts can require extensive processing, to
incorporate output from ensembles of different climate models and
different simulations of each model. This processing can be particularly
extensive when identifying and characterizing extreme events like heat
waves and frost day spells, as these must be processed from model output
with daily time steps and may use identification threholds based on
local distributions (e.g., 98\textsuperscript{th} percentile temperature
for that location). Here, we provide an overview of working with daily
climate model output data in R. Further, we present the
\pkg{futureheatwaves} package, which we developed to ease the process of
identifying, characterizing, and exploring multi-day extreme events in
climate model output. This package takes as input a directory of
projection files from one or many climate models, identifies all of a
specific type of extreme event (e.g., heat waves) for multiple cities
using a user-specified event definition, and writes an output directory
of files with these extreme events and their characteristics (e.g.,
length, absolute and relative temperatures). The \pkg{futureheatwaves}
package then allows the user to create custom functions and apply them
across all event files; this functionality can be used either to explore
extreme event patterns (e.g., average length and intensity across
different climate models) or to apply epidemiological estimates to
project health impacts under different scenarios.
}

@@

The \pkg{futureheatwaves} package automates the process of identifying
and characterizing multi-day extreme events across different ensemble
members of one or more climate models. A variety of different heat wave
definitions have been used to identify heat waves in a time series of
temperature data \citep{smith2013heat}, and the choice of heat wave
definitions can influence both health effect estimates
\citep{chen2015influence, kent2014heat} and projected heat wave trends
\citep{smith2013heat}. Further, other types of extreme events will be
defined differently than heat waves (for example, frost day spells may
be defined as one or more days with temperature at or below
32\textsuperscript{o}F / 0\textsuperscript{o}C). This package therefore
allows the user to create and use a custom the extreme event definition
used to identify events in the climate model output. Finally, this
package allows the user to explore the extreme events identified in the
climate model output with a function that can take a user-defined R
function and apply it across all extreme event files generated for the
separate ensemble members.

@@

We created the \pkg{futureheatwaves} package to aid in identifying,
characterizing, and exploring multi-day extreme events in daily climate
model output data. Figure \ref{fig:overview} gives an overview of the
two primary functions of the \pkg{futureheatwaves} package. First, the
\code{gen\_hw\_set} function processes a directory of climate projection
files that are stored locally on the user's computer (Figure
\ref{fig:overview}, ``Climate projections''), to generate a list of all
extreme events in each projection, as well as over a dozen
characteristics of each identified extreme event (Table
\ref{tab:hwcharacteristics}). The extreme events are identified and
characterized at one or more study locations (e.g., cities), with study
locations specified by the user in an input file. The extreme events
identified for each ensemble member are output as separate files in a
directory specified by the user (Figure \ref{fig:overview}, ``Extreme
events datasets``).

Once the user creates these data frames of location-specific heat waves,
the \code{apply\_all\_models} function allows the user to apply custom
functions across all the extreme event data frames generated by
\code{gen\_hw\_set} to summaries of extreme events across all climate
models and ensemble members (Figure \ref{fig:overview}, right). This
functionality can be used to generate summary statistics (e.g.,
determine average heat wave length or total frost days), or can be used
to apply more complex functions (e.g., apply epidemiologic effect
estimates across the heat waves to generate health impact estimates).

@@

@@

We have included data files in the package to serve as example files, so
users can try this package before applying it to their own directory of
climate projection files. This example data comes from two climate
models that are a part of CMIP5: (1) the model of the Beijing Climate
Center, China Meteorological Administration (BCC)
\citep{xin2013introduction} and (2) the National Center for Atmospheric
Research's (NCAR's) Community Climate System Model, version 4 (CCSM4)
\citep{gent2011community}. We include one ensemble member from BCC
(r1i1p1) and two from CCSM (r1i1p1 and r2i1p1).

To ensure that the size of this example data is reasonably small, we
have only included projection data for grid points from these climate
models that are near five U.S. east coast cities: New York, NY;
Philadelphia, PA; Newark, NJ; Baltimore, MD, and Providence, RI.
Further, to keep the file sizes reasonably small, the historical
projections range over the years 1990 to 1999, while the future
projections are limited to 2060 to 2079. Users' applications of this
package will likely use directories with many more climate model
ensemble members and more locations; however, the operation of the
package is the same for this smaller example application as it would be
for a much larger application.

Once the \pkg{futureheatwaves} package is installed and loaded, the user
can find the local location of these files using R's \code{system.file}
function. In the later sections of this article, we show how to use the
package functions with these example files as inputs.

When using this package with data other than this example data, the
climate model output data will require some pre-processing. Data will
need to be saved in a specific format, with files stored in a specific
directory structure. This preprocessing can be done using some of the R
packages described earlier for working with CMIP5 data in R. Full
details of the required file and directory structure are provided in the
package's ``futureheatwaves'' vignette, while tips on conducting this
processing starting from CMIP5 netCDF files are given in the
``starting\_from\_netcdf'' vignette.

\subsection{Basic example of identifying and characterizing extreme
events}\label{basic-example-of-identifying-and-characterizing-extreme-events}

Once climate model output files are set up as specified in the
``futureheatwaves'' package vignette, the package can process them to
identify and characterize heat waves in each ensemble member's
projection for each location using the \code{gen\_hw\_set} function. For
example, to process the example climate model output data included with
the package, the user can run:

@@

This code first identifies and saves as objects the path names on the
user's computer of the example climate projections directory
(\code{projection\_dir\_location}) and the file of study locations
(\code{city\_file\_location}). The \code{gen\_hw\_set} function
processes this example input and creates a new directory,
\file{example\_results}, with files of identified and characterized heat
waves, in the user's current working directory. In this example code,
the processing is done using default values for the event definition,
years for which to generate the heat wave data sets, etc. How and why to
customize these choices are explained later in the text. Function
arguments (e.g., \code{dataDirectories}, \code{tasFilenames}) are used
to specify the format of the data and the directory structure.

Once the function has completed running, results will be written locally
to the directory specified by the \code{out} argument of
\code{gen\_hw\_set}. This directory will include files with some basic
information about the climate models and the closest grid points of each
climate model to each location, as well as a directory with files of
identified and classified extreme events for each ensemble member,
including all characteristics in Table \ref{tab:hwcharacteristics}.

@@

By default, the package identifies extreme events in climate model
output data using a specific definition for heat waves:

\begin{quote}
A \dfn{heat wave} is two or more days at or above a city-specific
threshold temperature, with the threshold determined as the
98\textsuperscript{th} percentile of year-round temperature in the city
during some reference period (by default, 1990--1999).
\end{quote}

\noindent However, the user can extensively customize the definition
used to identify extreme events. Users can write a custom R function
with either a different heat wave definition (see \citet{smith2013heat}
and \citet{kent2014heat} for listings of definitions used in scientific
studies) or with a definition appropriate for a different type of
extreme event (e.g., one or more days at or below 32\textsuperscript{o}F
/ 0\textsuperscript{o}C for frost day spells). For heat wave
identification, researchers might want to use a specific definition, for
example, because it matches the definition used by local health
officials to declare heat wave warnings or, in the case of health impact
assessments, to match with a definition used in an epidemiological
study. For other extreme events, a heat wave definition likely will not
be applicable and so a customized definition is necessary.

Three components of the extreme event definition can be easily
customized in the \code{gen\_hw\_set} function call, without creating a
new R function to use to identify heat waves. First, often users wish to
define an extreme event based on conditions that are rare in the study
location \citep{IPCCch1}. In this case, the percentile of the variable
of interest required for an extreme event can be changed using the
\code{probThreshold} option in \code{gen\_hw\_set}. For example, this
option could be used to identify heat waves based on the
99\textsuperscript{th} percentile, rather than the
98\textsuperscript{th} percentile, of a city's year-round temperature.
Second, the user can change the number of days used in the heat wave
definition using the \code{numDays} argument in the \code{gen\_hw\_set}
function (the default is \samp{numDays = 2}, corresponding to two or
more days). Combined, these two customization choices allow the user to
identify heat waves using many of the heat wave definitions used in
previous climate and health research-- for example, 9 of the 16 heat
wave definitions used in \citet{kent2014heat} (e.g., {[}some examples of
the 9 that could be fit{]}) could be fit using different combinations of
these two options for specifying threshold percentile and number of
days. Third, some extreme events like cold waves and frost day spells
are defined as a certain number of days below, rather than above, a
threshold. While the default is to identify events by searching for days
above a threshold, this behavior can be changed with the
\code{above\_threshold = FALSE} argument in the
\code{gen\_hw\_set function}.

Beyond these simpler options, the customization of the event definition
is even more extensive as one has the option of writing a custom R
function. This functionality allows the user to use heat wave
definitions that either require a number of days above an absolute
threshold (e.g., maximum temperature of \(\ge\) 95\textsuperscript{o}F
for \(\ge1\) day \citet{kent2014heat, tan2007heat}; minimum temperature
\(\le\) 0\textsuperscript{o}C for \(\ge1\) day for frost day spells) or
that require a combination of thresholds to be met (e.g., maximum daily
temperature one threshold every day of the heat wave and above a higher
threshold for a certain number of days;
\citet{kent2014heat, peng2011toward}). To use a customized event
definition, the user can write and load an R function that implements
the definition. The custom function is passed to the \code{gen\_hw\_set}
function using the \code{IDheatwavesFunction} argument. To work
correctly, this custom function must allow only specific inputs and
generate only specific outputs. Details about this required structure
are provided in the \pkg{futureheatwaves} package vignette. To increase
processing speed when identifying extreme events, we coded parts of the
default event definition function in C++ and synced it with R using the
\CRANpkg{Rcpp} package \citep{Rcpp}. Users should consider a similar
strategy for custom heat wave definitions, especially if processing a
large number of climate projection files.

Finally, extreme events are often based on conditions that are rare for
a specific location using location-specific relative thresholds. These
thresholds are often defined for climate impact studies based on a
variable's distribution at that location in present-day or historical
data. However, it can be interesting to explore trends in extreme events
under climate change if extreme events are identified based on variable
distributions during the projection period, or another future period. It
is possible to specify the time period to use when determining a
relative threshold for an event definition using the
\code{thresholdBoundaries} argument in the function \code{gen\_hw\_set}.

@@

Once the user has created a directory of characterized event files for
each ensemble member (``Extreme event data sets'', Figure
\ref{fig:overview}), he or she can explore the results using the
\code{apply\_all\_models} function. This function allows the user to
apply custom R functions across all extreme event data frames created by
the \code{gen\_hw\_sets} call. The user can apply any R function that
follows certain standards in accepting input and returning output. Full
details on these standards are given in the \pkg{futureheatwaves}
package vignette.

As an example, if the user wanted to calculate the average temperature
of the heat waves identified for each ensemble member, he or she could
write a simple function:

@@

\noindent This function could then be applied across all extreme event
data sets output by \code{gen\_hw\_set} using the
\code{apply\_all\_models} function. For example, to apply this function
to all the example output results that come with the package, the user
could run:

@@

This output gives the results (\code{value} column) of running the
custom function for each ensemble member of each climate model. Note
that the location of the directory with the heat wave data frames must
be specified using the \code{out} argument when calling
\code{apply\_all\_models}. Typically, this will be the directory path
for the directory specified with the \code{out} argument in
\code{gen\_hw\_set}.

Location-specific results can be generated using the
\code{city\_specific} argument in \code{apply\_all\_models}:

@@

The \code{apply\_all\_models} function can also be used to project the
health impacts of heat waves. A simple estimate of excess deaths
associated with heat wave risk in a community can be calculated as
\citep{peng2011toward}:

\begin{equation}
E_c = B_c * (RR - 1) * L_c 
\label{eq:healthimpacts}
\end{equation}

@@

\begin{itemize}
\item $E_c$ is the total number of excess deaths in community $c$;
\item $B_c$ is the baseline average daily mortality in community $c$; 
\item $RR$ is the relative risk of mortality associated with a heat wave day; and 
\item $L_c$ is the total number of heat wave days in community $c$ over the study period.
\end{itemize}

This impact assessment equation can be translated into a function that
merges each projection's heat wave data frame with a data frame of
community-specific baseline mortality rates (\(B_c\)), calculates
equation \ref{eq:healthimpacts} for each heat wave, and then sums up the
total excess deaths across all heat waves. In this example, we have set
the default relative risk for a heat wave as 1.032, based on the
estimated added heat wave relative risk in 107 U.S. communities for heat
waves defined as two or more days at or above a community's
98\textsuperscript{th} percentile temperature
\citep{anderson2009weather}.

\begin{verbatim}
excess_deaths <- function(hw_datafr, base_mortality, RR = 1.032){
        hw_datafr <- dplyr::left_join(hw_datafr, base_mortality,
                                      by = "city") %>%
                dplyr::mutate(excess_deaths = base_mort * (RR - 1) * length)
        out <- sum(hw_datafr$excess_deaths)
        out <- round(out)
        return(out)
}
\end{verbatim}

\noindent Once defined in R, this function can be applied across all
heat waves from all climate models' ensemble members, provided that you
have a data frame called \code{city\_mortality} with columns with each
community's identifier (\code{city}) and baseline mortality rate
(\code{base\_mort}), using the call (this example uses hypothetical
baseline mortality rates for the cities and requires the \CRANpkg{dplyr}
package \citep{dplyr}):

\begin{Schunk}
\begin{Sinput}
library(dplyr)
city_mortality <- data.frame(city = c("balt", "nwk", "ny", "phil", "prov"),
                             base_mort = c(3, 2.5, 35, 6, 2.5))
apply_all_models(out = out, FUN = excess_deaths, base_mortality = city_mortality)
\end{Sinput}
\begin{Soutput}
#>   model ensemble value
#> 1  bcc1        1   930
#> 2  ccsm        1   968
#> 3  ccsm        2   967
\end{Soutput}
\end{Schunk}

\noindent While this example impact analysis is simplistic, it provides
an example of how the \code{apply\_all\_models} function can be used in
conjunction with epidemiologic results to estimate the potential health
impacts of extreme events with climate change.

@@

The heat wave data sets characterize heat wave in several ways that are
based on relative temperature (for example, the
\texttt{mean.temp.quantile}, \texttt{max.temp.quantile}, and
\texttt{min.temp.quantile} columns described in Table
\ref{tab:hwcharacteristics}). These characteristics are measured for
each of the heat waves identified by the \texttt{gen\_hw\_set} function
by taking the absolute temperature of the heat wave (e.g., average
temperature during the heat wave is 90\textsuperscript{o}F) and
comparing it to the heat wave location's typical temperature
distribution. This process generates relative measures of how intense
the heat wave is compared to what is normal in that location (e.g.,
90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of
year-round temperatures in the location).

The values of these relative heat wave characteristics will depend on
what time period is used to determine the location's typical temperature
distribution. For example, the same heat wave would likely have much
higher values for these relative characteristics if the temperature
during the heat wave is compared to typical temperatures in the present
day for that location versus if the temperature is compared to typical
temperatures much later in the 21\textsuperscript{st} century for the
same location.

@@

\begin{figure}
\begin{center}
\includegraphics[width = 0.9\textwidth]{ExampleLeaflet}
\end{center}
\caption{Snapshot of an example interactive map created using the \code{map\_grid\_leaflet} showing the locations of study cities and their matching climate model grid points for the BCC climate model example data included with \pkg{futureheatwaves}. The lines on the map connect each climate model grid point to the study location(s) for which that grid point was used. The interactive maps include pop-ups with city identiers; one is shown open in this snapshot as an example. From this map, you can see that the climate model grid point closest to New York City for this climate model is over the Atlantic Ocean.}
\label{fig:gridmap}
\end{figure}
